HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-10-14 docs/HLL_exp_bad/main.tex
tweaks

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if the HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We discuss and benchmark a suitable choice for the required VIL-PRF.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. We model this by assuming the adversary has access to a large set ${\cal X}$ of potential inputs from which it can select items to insert into the HLL. Our adversary's objective, then, is to insert as many items from ${\cal X}$ as possible whilst keeping the HLL cardinality estimate as low as possible.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL sketch but can access a shadow copy of the HLL sketch via its API. It can use the API to insert elements into this shadow HLL sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. 
%Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

Now we turn to the case where the HLL under attack is not initially empty, but instead contains some honestly inserted items. As usual, the adversary's goal is to insert as many items as possible whilst increasing the cardinality estimate as little as possible. We focus on scenarios S2 and S4, and make brief remarks on S1.

@@

\subsubsection{Initially Non-empty HLL Sketch, Scenario S2:}\label{sec:nonemptyHLLS2}
The adversary knows $h$ so it selects items $x$ from ${\cal X}$ for which the $\ell$ rightmost bits of $h(x)$ start with a ``1'' bit. In this case, since the relevant bucket $j$ already has $M[j] \ge 1$, the bucket's counter is not updated, and the cardinality estimate of the HLL remains unchanged.

The adversary then expects to be able to insert half of the items from ${\cal X}$ into the HLL, without increasing the HLL cardinality estimate at all.

@@

However, it is clear that any adversary in the ideal world is reduced to operating in the setting used in the average case analysis of HLL (wherein every insertion query results in a random value $y$ being used to update the HLL). Thus, ${\cal A}$ can have no advantage when it comes to influencing the HLL cardinality estimate. It follows that HLL, when instantiated with PRF $F$, is secure against attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well). We will conclude by discussing how $F$ can be instantiated.

@@

We consider adversary ${\cal A}$ to be an algorithm running against a challenger ${\cal C}$ in a security game. The game begins with ${\cal C}$ selecting a bit $b$ uniformly at random. Adversary ${\cal A}$ interacts with up to four oracles which are handled by ${\cal C}$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\cal A}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). For $i=1, 2, 3$, we define ${\cal A}$'s S$i$-Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\cal A}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\cal A}$ can interact with are as follows:
\begin{itemize}
\item \texttt{init}: this oracle performs initialisation of the HLL (and selection of the PRF key when $h$ is replaced by a PRF $F$); this oracle is available in all of S2, S3, S4, but can only be called once as ${\cal A}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {\cal D}$, the oracle inserts value $x$ into the HLL. This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\cal A}$ to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\cal A}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate. We allow multiple queries to this oracle and make it available to ${\cal A}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We conclude by reiterating our central point: we have shown that no efficient S4 adversary ${\cal A}$ can tell that it is playing against HLL implemented using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\cal A}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. 
%It does not matter that ${\cal A}$ can adaptively choose its inputs $x$ in \texttt{insert($\cdot$)} queries, nor that it can access \texttt{read} and \texttt{corrupt} queries. For each \texttt{insert($\cdot$)} query, a fresh value $x$ leads to a new random $y$ being generated and used to update a bucket, as in the average case analysis of HLL. 
%The \texttt{read} and \texttt{corrupt} queries do leak information about $F$'s outputs to the adversary, but such outputs are indistinguishable from random under the assumption that $F$ is a PRF.

@@

The above construction requires a PRF $F: \{0,1\}^k \times {\cal D} \rightarrow  \{0,1\}^{n+\ell}$. Here, the set ${\cal D}$ represents the space of possible inputs to the HLL which can be modelled by a set of bit-strings of some maximum length. This mean that $F$ needs to be Variable Input Length PRF (VIL-PRF).

@@

\todo{Add a sentence on SipHash-2-4 performance here. Microbenchmark on a short input suffices.}

>>>>

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We discuss and benchmark a suitable choice for the required VIL-PRF.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. We model this by assuming the adversary has access to a large set ${\cal X}$ of potential inputs from which it can select items to insert into the HLL sketch. Our adversary's objective, then, is to insert as many items from ${\cal X}$ as possible whilst keeping the HLL cardinality estimate as low as possible.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL sketch but can access a shadow copy of the HLL sketch via its API. It can use the API to insert elements into this shadow HLL sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL sketch and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. 
%Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

Now we turn to the case where the HLL sketch under attack is not initially empty, but instead contains some honestly inserted items. As usual, the adversary's goal is to insert as many items as possible whilst increasing the cardinality estimate as little as possible. We focus on scenarios S2 and S4, and make brief remarks on S1.

@@

\subsubsection{Initially Non-empty HLL Sketch, Scenario S2:}\label{sec:nonemptyHLLS2}
The adversary knows $h$ so it selects items $x$ from ${\cal X}$ for which the $\ell$ rightmost bits of $h(x)$ start with a ``1'' bit. In this case, since the relevant bucket $j$ already has $M[j] \ge 1$, the bucket's counter is not updated, and the cardinality estimate of the HLL sketch remains unchanged.

The adversary then expects to be able to insert half of the items from ${\cal X}$ into the HLL sketch, without increasing the HLL cardinality estimate at all.

@@

However, it is clear that any adversary in the ideal world is reduced to operating in the setting used in the average case analysis of HLL (wherein every insertion query results in a random value $y$ being used to update the HLL sketch). Thus, ${\cal A}$ can have no advantage when it comes to influencing the HLL cardinality estimate. It follows that HLL, when instantiated with PRF $F$, is secure against attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well). We will conclude by discussing how $F$ can be instantiated.

@@

We consider adversary ${\cal A}$ to be an algorithm running against a challenger ${\cal C}$ in a security game. The game begins with ${\cal C}$ selecting a bit $b$ uniformly at random. Adversary ${\cal A}$ interacts with up to four oracles which are handled by ${\cal C}$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\cal A}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). For $i=1, 2, 3$, we define ${\cal A}$'s S$i$-Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\cal A}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\cal A}$ can interact with are as follows:
\begin{itemize}
\item \texttt{init}: this oracle performs initialisation of the HLL sketch (and selection of the PRF key when $h$ is replaced by a PRF $F$); this oracle is available in all of S2, S3, S4, but can only be called once as ${\cal A}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {\cal D}$, the oracle inserts value $x$ into the HLL sketch. This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\cal A}$ to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\cal A}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate. We allow multiple queries to this oracle and make it available to ${\cal A}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We conclude by reiterating our central point: we have shown that no efficient S4 adversary ${\cal A}$ can tell that it is playing against HLL implemented using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\cal A}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. 
It does not matter that ${\cal A}$ can adaptively choose its inputs $x$ in \texttt{insert($\cdot$)} queries, nor that it can access \texttt{read} and \texttt{corrupt} queries. For each \texttt{insert($\cdot$)} query, a fresh value $x$ leads to a new random $y$ being generated and used to update a bucket, as in the average case analysis of HLL. The \texttt{read} and \texttt{corrupt} queries do leak information about $F$'s outputs to the adversary, but such outputs are indistinguishable from random under the assumption that $F$ is a PRF.

@@

The above construction requires a PRF $F: \{0,1\}^k \times {\cal D} \rightarrow  \{0,1\}^{n+\ell}$. Here, the set ${\cal D}$ represents the space of possible inputs to HLL which can be modelled by a set of bit-strings of some maximum length. This mean that $F$ needs to be Variable Input Length PRF (VIL-PRF).

@@

For example, our microbenchmarks created on an Intel i7 CPU show that SipHash-2-4 is as fast as the 128-bit version of MurmurHash3 and only slightly slower than the original 32-bit version.
